Deciding Bitvector Termination with Program Synthesis

### 2014-07-06 papers/termination/paper.tex
CBMC synth
formalisation

\section{Motivational Examples} \label{sec:motivation}
- see phase change programs.

@@

where $\vec{P}$ ranges over sets, while $\vec{x}$ and $\vec{y}$ range over ground terms.
Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to \mathbb{B}$ can be viewed as a specification function
that  returns true iff the functions $\vec{P}$ compute appropriate outputs
when fed the inputs $\vec{x}$ and $\vec{y}$.  
\end{definition}

Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
%Solving the synthesis problem 
amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
for all $\vec{y}$. 
If a pair $(\vec{P_0}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P_0}, \vec{x}) \models \sigma$.
For the remainder of the presentation, we drop the $\vec{x}$ and instead just write $x$, with the understanding
that all variables range over vectors.

@@

\noindent {\bf Environment abstractions.}
For a loop $L=(G,T)$, the formulae {\bf [CT]} and {\bf [CNT]} from Definition~\ref{def:ct} and Definition~\ref{def:cnt} are in terms of 
$T_o$ and $T_u$, which over-approximate and under-approximate the transition relation $T$, respectively. Moreover, in formula {\bf [CT]},  
$I$ over-approximates the set of states that $L$ may begin executing in.
The reason for using over- and under-approximations [instead of exact relations] when formulating the (non-)termination of an isolated loop 
is so that we can use the same formulae regardless of the environment the loop is executing in, or 
the complexity of its body, e.g. if it contains nested loops.

All we need to do in order to use {\bf [CT]} and {\bf [CNT]} in any context %environment and regardless of the complexity of the loops's body 
is generate constraints abstracting 
the environment and the loop's body: for termination, we need over-approximating abstractions, whereas for non-termination 
we need under-approximations. 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through examples.\\

@@

\noindent {\bf Over-approximating the environment for termination.}
%% While proving that loop $L$ terminates, we create constraints on the environment in the
%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
%% In the presence of nested loops, 
%% for each loop $L'$ nested in the body
%% of $L$, we must generate constraints saying that $T_o$ overapproximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
%The details of how to create these constraints are standard, so
%rather than presenting the algorithm for generating the constraints, we direct the
%reader's attention to Figure~\ref{fig:environment-model}.  
%
The example in Figure~\ref{fig:environment-model} contains three loops. %, which we will number 1 to 3 according to their program order.
Let us assume that we have already proven the termination of $L_1$ and $L_3$, and we are now trying to prove that $L_2$ terminates, which we encode
as the conjunction of constraints on the right side of the figure:
\begin{itemize}
\item $W_1$, $W_2$ and $W_3$ are (over-approximating) inductive invariants for $L_1$, $L_2$ and $L_3$, respectively.
\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R$ is a ranking function for $L_2$.
 \end{itemize}

@@

\noindent {\bf Under-approximating the environment for non-termination.}
Dual to termination, when attempting to prove non-termination, we must under-approximate the environment and the loop body's and apply 
formula {\bf [CNT]}. However, there is one more constraint that we must obey when doing so: we must stay inside the synthesis fragment! 
This can be achieved for the environment as its under-approximation $I$ and the loop under consideration are at the same level of nestedness and do not therefore 
require another quantifier alternation.

However, when dealing with nested loops, computing the under-approximation of the loop's body 
requires an extra quantifier alternation. The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation. This situation is similar to the one encountered when proving non-termination of non-deterministic loops. 
To go around this limitation, we treat nested loops as a special case and give a separate formula for them in Figure~\ref{fig:nonterm-nested}. If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
\begin{itemize}
\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
\item If $L_2$ is non-terminating, then $N_2$ is its recurrence set.
\end{itemize}

@@

>>>>

\section{Motivational Examples} \label{sec:motivation}
%- see phase change programs.

@@

where $\vec{P}$ ranges over sets, while $\vec{x}$ and $\vec{y}$ range over ground terms.
Function $\sigma: (X^n \times Y^m \to Z^k) \times X^n \times Y^m  \to \mathbb{B}$ can be viewed as a specification function
that  returns true iff the functions $\vec{P}$ compute appropriate outputs
when fed the inputs $\vec{x}$ and $\vec{y}$.  
\end{definition}
%
Checking satisfiability of a formula in the synthesis fragment corresponds to program synthesis and
%Solving the synthesis problem 
amounts to finding witnesses for the unknowns $\vec{P}$ and $\vec{x}$ that meet the specification
for all $\vec{y}$. 
If a pair $(\vec{P_0}, \vec{x})$ is a solution to the synthesis problem, then we write $(\vec{P_0}, \vec{x}) \models \sigma$.
For the remainder of the presentation, we drop the $\vec{x}$ and instead just write $x$, with the understanding
that all variables range over vectors.

@@

@@

\noindent {\bf Over-approximating the environment for termination.}
%% While proving that loop $L$ terminates, we create constraints on the environment in the
%% form of the predicate $I$ and the over-approximating summary relation $T_o$:
%% In the presence of nested loops, 
%% for each loop $L'$ nested in the body
%% of $L$, we must generate constraints saying that $T_o$ overapproximates $L'$, i.e. it is an over-approximation of the transitive closure of $L'$'s transition relation.
%% Similarly, for each loop $L''$ appearing before $L$ in the program's control flow, we
%% must add constraints saying that $I$ over-approximates the states reachable via $L''$.
%The details of how to create these constraints are standard, so
%rather than presenting the algorithm for generating the constraints, we direct the
%reader's attention to Figure~\ref{fig:environment-model}.  
%
%\noindent {\bf Environment abstractions.}
For a loop $L=(G,T)$, the formulae {\bf [CT]} %and {\bf [CNT]} 
from Definition~\ref{def:ct} 
%and Definition~\ref{def:cnt} 
is in terms of 
$T_o$ and $I$, which over-approximate the transition relation $T$ and the set of states that $L$ may begin executing in, respectively. 
The reason for using over-approximations [instead of exact relations] %when formulating the (non-)termination of an isolated loop 
is so that we can use the same formulae regardless of the environment the loop is executing in, or 
the complexity of its body, e.g. if it contains nested loops.

All we need to do in order to use {\bf [CT]} in any context %environment and regardless of the complexity of the loops's body 
is generate constraints over-approximating %abstracting 
the environment and the loop's body. %: for termination, we need over-approximating abstractions, whereas for non-termination 
%we need under-approximations. 
As the generation of such constraints is standard and covered by several other works \cite{DBLP:conf/pldi/GrebenshchikovLPR12,DBLP:conf/pldi/GulwaniSV08}, 
we will not provide the full algorithm, but rather illustrate it through the example in Figure~\ref{fig:environment-model}.
Let us assume that we have already proven the termination of loops $L_1$ and $L_3$, and we are now trying to prove that $L_2$ terminates, which we encode
as the conjunction of constraints on the right side of the figure:
\begin{itemize}
\item $W_1$, $W_2$ and $W_3$ are (over-approximating) inductive invariants for $L_1$, $L_2$ and $L_3$, respectively.
\item $S$ is a summary of $L_3$ that over-approximates the transitive closure of its transition relation.
\item If the formula is satisfiable, then $R$ is a ranking function for $L_2$.
 \end{itemize}

@@

\noindent {\bf Under-approximating the environment for non-termination.}
Dual to termination, when proving non-termination, we need to under-approximate the environment and the loop's body and apply
formula {\bf [CNT]}. %As opposed to over-approximations 
%this is not as straightforward as when computing over-approximations in the termination case described above.
Recall that there is one more constraint that we must obey when doing so: we must stay inside the synthesis fragment! 
This becomes more challenging now then when dealing with over-approximations due to the existential nature of under-approximations: they require an additional existential quantifier, which might result in an additional quantifier alternation, i.e. $\exists \forall \exists$.

This issue does not arise when abstracting the environment as its under-approximation and the loop under consideration make use of existential quantification at the same nestedness level. However, when dealing with nested loops, we need to compute the under-approximation of the inner loop, which requires a nested existential quantifier, 
resulting in the $\exists \forall \exists$ alternation. 
%% computing the under-approximation of the loop's body requires an extra quantifier alternation. 
%% The origin of the problem comes from the fact that an under-approximation is an existential property and therefore
%% requires an extra existential quantifier, resulting in the $\exists \forall \exists$ alternation. This situation is similar to the one encountered when proving non-termination of non-deterministic loops. 
To go around this limitation, we treat nested loops as a special case and give a separate formula for them in Figure~\ref{fig:nonterm-nested}. If the formula on the right of the figure is satisfiable, then $L_1$ is non-terminating as witnessed by the recurrence set $N_1$ and
the initial state $x_0$ in which the program begins executing. There are two possible scenarios for $L_2$'s termination:
\begin{itemize}
\item If $L_2$ is terminating, then $N_2$ is an inductive invariant that reestablished $N_1$ after $L_2$ stops executing: $\lnot G_2(x) \wedge N_2(x) \wedge P_2(x,x') \rightarrow N_1(x') $. 
\item If $L_2$ is non-terminating, then $N_2$ is its recurrence set.
\end{itemize}

@@

\noindent{\bf Symbolic Bounded Model Checking} Another complete method for generating
candidates is to simply use BMC on the {\sc synth.c} program.  As with explicit
search, we must progressively increase the length of the solution we search for
in order to get a complete search procedure.
\\
